\textbf{The Heart can Lie: A Preliminary Investigation of the Role of Interoception and Theory of Mind in Deception}

### 2024-07-10 paper/manuscript.Rmd
R2

Lying - the intentional attempt at instilling a false belief in others [@sip2012if] - is a prevalent phenomenon carrying potentially important consequences. Interestingly, evidence suggests that the successful detection of a lying attempt depends more on the ability of the liar, than on the performance of the lie detector [@verigin2019lie; @bond2008individual; @levine2011sender]. However, with most of the deception literature focused on deception detection [@viji2022history; @sternglanz2019review; @masip2017deception], the factors contributing to one's ability to lie remain unclear. Nevertheless, some findings suggest a relationship between the propensity to tell lies, and traits that characterize the socially malevolent profile known as the Dark Triad [@paulhus2002dark], such as narcissism [@zvi2018correlates] and psychopathy [@rassin2023psychopathy]. While often conceptualized to be immoral and unconscionable, lying is ubiquitous in everyday life, and being able to lie skillfully can sometimes facilitate interpersonal relationships, helping us avoid conflict or causing emotional harm to others [@levine2022prosocial]. In fact, recent research shows that certain forms of deception, such as prosocial lies (i.e., false statements told to benefit others; @levine2022prosocial), can increase trust [@levine2015prosocial]. Moreover, individuals who told altruistic lies were perceived as more benevolent than those who were honest [@levine2014liars].

@@

The present study examined the contribution of ToM and interoception on our ability to lie using a directed lie paradigm with two conditions ("Interrogation" and "Polygraph") designed to enhance each of the two mechanisms. Interestingly, we found that when participants' responses were perceived to be evaluated by a person (the interrogation condition), instead of (fake) lie detection machine (the polygraph condition), their response time for both lies and truths were faster, and their heart rate was elevated. Although the condition did not impact the subjective confidence that participants had in their answers, the pattern of results suggests that believing one's response is being evaluated by a person, instead of a machine, could induce more fear, consequently speeding up the response and increasing the physiological arousal [@aylward2017impact]. Alternatively, the slower response in the polygraph condition could be explained by the established attentional switching hypothesis, which posits that an increase in attention towards internal signals and managing one's emotional reaction would confer less cognitive resources available, thereby resulting in individuals taking a longer time to respond [@arnold2019interoception; @hanania2010selective]. While the impacts of external settings on individuals' responses warrant further investigation, the results highlight how physiological responses can be easily confounded by other factors (e.g., presence or absence of a stimuli), independent of whether one is lying or telling the truth. By extension, our study concurs with the controversial discourse surrounding the use of physiological measures in deception research [@rosky2013f; @oviatt2018handbook].

>>>>

Lying - the intentional attempt at instilling a false belief in others [@sip2012if] - is a prevalent phenomenon carrying potentially important consequences. Interestingly, evidence suggests that the successful detection of a lying attempt depends more on the ability of the liar, than on the performance of the lie detector [@verigin2019lie; @bond2008individual; @levine2011sender]. However, with most of the deception literature focused on deception detection [@viji2022history; @sternglanz2019review; @masip2017deception], the factors contributing to one's ability to lie remain unclear. Nevertheless, some findings suggest a relationship between the propensity to tell lies, and traits that characterize the socially malevolent profile known as the Dark Triad [@paulhus2002dark], such as narcissism [@zvi2018correlates] and psychopathy [@rassin2023psychopathy]. While often conceptualized to be immoral and unconscionable, lying is ubiquitous in everyday life, and being able to lie skillfully can sometimes facilitate interpersonal relationships, helping us avoid conflict or causing emotional harm to others [@levine2022prosocial]. In fact, recent research shows that certain forms of deception, such as prosocial lies [i.e., false statements told to benefit others, @levine2022prosocial], can increase trust [@levine2015prosocial]. Moreover, individuals who told altruistic lies were perceived as more benevolent than those who were honest [@levine2014liars].

@@

The present study examined the contribution of ToM and interoception on our ability to lie using a directed lie paradigm with two conditions ("Interrogation" and "Polygraph") designed to enhance each of the two mechanisms. Interestingly, we found that when participants' responses were perceived to be evaluated by a person (the interrogation condition), instead of the lie detection machine (the polygraph condition), their response time for both lies and truths were faster, and their heart rate was elevated. Although the condition did not impact the subjective confidence that participants had in their answers, the pattern of results suggests that believing one's response is being evaluated by a person, instead of a machine, could induce more fear, consequently speeding up the response and increasing the physiological arousal [@aylward2017impact]. Alternatively, the slower response in the polygraph condition could be explained by the established attentional switching hypothesis, which posits that an increase in attention towards internal signals and managing one's emotional reaction would confer less cognitive resources available, thereby resulting in individuals taking a longer time to respond [@arnold2019interoception; @hanania2010selective]. While the impacts of external settings on individuals' responses warrant further investigation, the results highlight how physiological responses can be easily confounded by other factors, independent of whether one is lying or telling the truth. For instance, the presence or absence of the "interrogator", or the saliency of the moral nature of the task [e.g. @peleg2019lie, argues that the polygraph test alone also acts as a "moral reminder", framing the possibility that physiological arousal in a polygraph context might be partially a reflection of individuals’ attention directed to their own moral standards]. By extension, our study concurs with the controversial discourse surrounding the use of physiological measures in deception research [@rosky2013f; @oviatt2018handbook].

### 2024-07-10 paper/manuscript.tex
R2

Lying - the intentional attempt at instilling a false belief in others (Sip et al., 2012) - is a prevalent phenomenon carrying potentially important consequences. Interestingly, evidence suggests that the successful detection of a lying attempt depends more on the ability of the liar, than on the performance of the lie detector (Bond Jr \& DePaulo, 2008; T. R. Levine et al., 2011; Verigin et al., 2019). However, with most of the deception literature focused on deception detection (Masip, 2017; Sternglanz et al., 2019; Viji et al., 2022), the factors contributing to one's ability to lie remain unclear. Nevertheless, some findings suggest a relationship between the propensity to tell lies, and traits that characterize the socially malevolent profile known as the Dark Triad (Paulhus \& Williams, 2002), such as narcissism (Zvi \& Elaad, 2018) and psychopathy (Rassin et al., 2023). While often conceptualized to be immoral and unconscionable, lying is ubiquitous in everyday life, and being able to lie skillfully can sometimes facilitate interpersonal relationships, helping us avoid conflict or causing emotional harm to others (E. E. Levine \& Lupoli, 2022). In fact, recent research shows that certain forms of deception, such as prosocial lies (i.e., false statements told to benefit others; E. E. Levine and Lupoli (2022)), can increase trust (E. E. Levine \& Schweitzer, 2015). Moreover, individuals who told altruistic lies were perceived as more benevolent than those who were honest (E. E. Levine \& Schweitzer, 2014).

@@

The present study examined the contribution of ToM and interoception on our ability to lie using a directed lie paradigm with two conditions (``Interrogation'' and ``Polygraph'') designed to enhance each of the two mechanisms. Interestingly, we found that when participants' responses were perceived to be evaluated by a person (the interrogation condition), instead of (fake) lie detection machine (the polygraph condition), their response time for both lies and truths were faster, and their heart rate was elevated. Although the condition did not impact the subjective confidence that participants had in their answers, the pattern of results suggests that believing one's response is being evaluated by a person, instead of a machine, could induce more fear, consequently speeding up the response and increasing the physiological arousal (Aylward et al., 2017). Alternatively, the slower response in the polygraph condition could be explained by the established attentional switching hypothesis, which posits that an increase in attention towards internal signals and managing one's emotional reaction would confer less cognitive resources available, thereby resulting in individuals taking a longer time to respond (Arnold et al., 2019; Hanania \& Smith, 2010). While the impacts of external settings on individuals' responses warrant further investigation, the results highlight how physiological responses can be easily confounded by other factors (e.g., presence or absence of a stimuli), independent of whether one is lying or telling the truth. By extension, our study concurs with the controversial discourse surrounding the use of physiological measures in deception research (Oviatt et al., 2018; Rosky, 2013).

>>>>

Lying - the intentional attempt at instilling a false belief in others (Sip et al., 2012) - is a prevalent phenomenon carrying potentially important consequences. Interestingly, evidence suggests that the successful detection of a lying attempt depends more on the ability of the liar, than on the performance of the lie detector (Bond Jr \& DePaulo, 2008; T. R. Levine et al., 2011; Verigin et al., 2019). However, with most of the deception literature focused on deception detection (Masip, 2017; Sternglanz et al., 2019; Viji et al., 2022), the factors contributing to one's ability to lie remain unclear. Nevertheless, some findings suggest a relationship between the propensity to tell lies, and traits that characterize the socially malevolent profile known as the Dark Triad (Paulhus \& Williams, 2002), such as narcissism (Zvi \& Elaad, 2018) and psychopathy (Rassin et al., 2023). While often conceptualized to be immoral and unconscionable, lying is ubiquitous in everyday life, and being able to lie skillfully can sometimes facilitate interpersonal relationships, helping us avoid conflict or causing emotional harm to others (E. E. Levine \& Lupoli, 2022). In fact, recent research shows that certain forms of deception, such as prosocial lies (i.e., false statements told to benefit others, E. E. Levine \& Lupoli, 2022), can increase trust (E. E. Levine \& Schweitzer, 2015). Moreover, individuals who told altruistic lies were perceived as more benevolent than those who were honest (E. E. Levine \& Schweitzer, 2014).

@@

The present study examined the contribution of ToM and interoception on our ability to lie using a directed lie paradigm with two conditions (``Interrogation'' and ``Polygraph'') designed to enhance each of the two mechanisms. Interestingly, we found that when participants' responses were perceived to be evaluated by a person (the interrogation condition), instead of the lie detection machine (the polygraph condition), their response time for both lies and truths were faster, and their heart rate was elevated. Although the condition did not impact the subjective confidence that participants had in their answers, the pattern of results suggests that believing one's response is being evaluated by a person, instead of a machine, could induce more fear, consequently speeding up the response and increasing the physiological arousal (Aylward et al., 2017). Alternatively, the slower response in the polygraph condition could be explained by the established attentional switching hypothesis, which posits that an increase in attention towards internal signals and managing one's emotional reaction would confer less cognitive resources available, thereby resulting in individuals taking a longer time to respond (Arnold et al., 2019; Hanania \& Smith, 2010). While the impacts of external settings on individuals' responses warrant further investigation, the results highlight how physiological responses can be easily confounded by other factors, independent of whether one is lying or telling the truth. For instance, the presence or absence of the ``interrogator'', or the saliency of the moral nature of the task (e.g. Peleg et al., 2019, argues that the polygraph test alone also acts as a ``moral reminder,'' framing the possibility that physiological arousal in a polygraph context might be partially a reflection of individuals' attention directed to their own moral standards). By extension, our study concurs with the controversial discourse surrounding the use of physiological measures in deception research (Oviatt et al., 2018; Rosky, 2013).
